Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2016-06-21 doc/invt.tex
small changes

To proceed, we first define an average\cite{
dama2014},
\begin{equation}
  \bar v(t)
  = \sum_{i = 1}^n \rho_i \, v_i(t)
\label{eq:vbar_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$.
%
This distribution $\pmb \rho$
will be used to measure the error,
and is independent of
the target distribution $\mathbf p$.
%
For example,
in well-tempered metadynamics,
$\pmb\rho$ is always flat,
while $\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
barducci2008, dama2014}.
%
Then, we expect the difference %between $v_i$ and $\bar v$,
%
\begin{equation}
  x_i \equiv v_i - \bar v
  %=
  %v_i - \sum_{j = 1}^n \rho_j \, v_j
  ,
\label{eq:x_def}
\end{equation}
%
to be small in the asymptotic regime.
%
It follows that
\begin{equation}
  \sum_{i = 1}^n \rho_i \, x_i = 0
  ,
\label{eq:rhox_sum}
\end{equation}
and
\begin{equation}
  \frac{ d \bar v (t) }
       { d t }
  =
  \alpha(t)
  \sum_{ r = 1 }^n \frac{ h_r(t) } { p_r }
  .
\label{eq:dvbardt}
\end{equation}
%
Expanding Eq. \eqref{eq:h_ave} in $x_i$ yields
%
\begin{align}
  \frac{
    \langle h_i(t) \rangle
  }
  {
    p_i
  }
  &\approx
  \frac{                        e^{- x_i} }
       { \sum_{ j = 1}^n p_j \, e^{- x_j} }
  \approx
  1 - x_i + \sum_{r = 1}^n p_r \, x_r
  ,
\label{eq:hdet_linearize}
\end{align}
%
where we have used $\sum_{j=1}^n p_j = 1$.
%
\note{For \eqref{eq:hdet_linearize}. Following the second step
$$
\langle h_i(t) \rangle
\approx
\frac{                  p_i (1 - x_i) }
     { \sum_{ r = 1 }^n p_r (1 - x_r) }
=
\frac{                      p_i (1 - x_i) }
     { 1 - \sum_{ r = 1 }^n p_r \, x_r }
.
$$
}%
From Eqs.
\eqref{eq:vt_diffeq},
\eqref{eq:h_split},
\eqref{eq:x_def}-\eqref{eq:hdet_linearize},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_WL}
\end{equation}
where
\begin{equation}
  \xi_i(t)
  =
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ r = 1 }^n
    \rho_r
    \frac{ \zeta_r(t) }
         {     p_r    }
  .
\label{eq:xi_def}
\end{equation}
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:dvbardt}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ r = 1 }^n
      \rho_r \frac{ h_r } { p_r }
  \right)
  \\
  &\stackrel{\mathrm{Eq.\; \eqref{eq:h_split}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ r = 1 }^n
      \rho_r \frac{ \langle h_r \rangle } { p_r }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:rhox_sum} \;and\; \eqref{eq:hdet_linearize}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

We define
the total error of the bias potential as
\begin{equation}
  \Err
  =
  \sum_{i = 1}^n \rho_i \, \langle x_i^2 \rangle
  =
  \sum_{i = 1}^n \rho_i \, \left\langle (v_i - \bar v)^2 \right\rangle
  .
\label{eq:error_sum}
\end{equation}
%
Our aim is to find the $\alpha(t)$
that minimizes the error $E$
at the end of the simulation period $T$.
%
It suffices to consider
a simple one-variable problem.

@@

To simplify the ensuing discussion,
we  limit ourselves to $\mathbf w$'s
that satisfy
the detailed balance condition,
%
\begin{equation}
  \rho_i \, w_{ij} = \rho_j \, w_{ji}
  ,
  \label{eq:w_detailedbalance}
\end{equation}
%
for the distribution $\pmb \rho$
introduced in Eq. \eqref{eq:vbar_def}.
%
It follows that
\begin{equation}
  \sum_{i = 1}^n \rho_i \, w_{ij}
  =
  \sum_{i = 1}^n \rho_j \, w_{ji}
  =
  \rho_j,
  \label{eq:w_balance}
\end{equation}
%
i.e., $\mathbf \rho$ is a left eigenvector of
$\mathbf w$ with eigenvalue $1$.

@@

In Fig. \ref{fig:sinc},
we give examples of the updating kernels
of the bandpass updating schemes,
compared to a Gaussian
with a similar width, $\sigma$,
\note{(cheng, the inset in figure two should be replotted to fix the dynamic range of the gray scale.)}
%
\begin{equation}
  \sigma
  \approx
  \frac
  {
    n
  }
  {
    \sqrt{ 2 \, \pi } \, g \, K
  }
  .
\label{eq:sigma_equiv}
\end{equation}
%
%where $g = 2$ for the periodic case
%or $g = 1$ otherwise.
%
The kernel in the non-periodic case is a zigzag
because of the term $(-1)^{n+K+l-1}$
in Eq. \eqref{eq:mu_sinc_refl}.
%
However, as shown in the inset, % of Fig. \ref{fig:sinc}(b),
the ruggedness disappears
in the updating matrix by Eq. \eqref{eq:w_band_refl},
as the above oscillatory term for $l = i - j$
is canceled by the term for either $l = i + j - 1$
or $l = i + j - 2 \, n - 1$
with the opposite sign.

@@

\note{(Cheng, we should give a  description of the test system)}
Our test system is a one-dimensional model system, in which
$i$ represents a non-periodic variable,
and $n = 100$.
%
Both the target and intrinsic distributions
are assumed to be flat:
$p^*_i = p_i = 1/n$.
%
%We wish to see how
%the schedule, $\alpha(t)$,
%affects the fluctuating error
%of the bias potential
%at the end of the simulation.
%
Initially,
we set the bias potential to zero,
$V_i \equiv 0$,
and equilibrate the system for $10^7$ steps
under a constant updating magnitude,
$\alpha_0 = 10^{-4}$.
%
After this,
we reset the origin of the time, $t$,
and start the testing schedule
for $T = 10^8$ steps.
%
We repeat the above process $1000$ times (unless specified otherwise)
and average the results.
%specified by either
%Eq. \eqref{eq:q_opt}
%or
%Eq. \eqref{eq:alpha_invtlambda}.

@@

\begin{table}[h]
  \caption{\label{tab:error_Gaussian}
    Error of the Gaussian updating scheme of $\sigma = 10$.
    $t_0$ is given by Eq. \eqref{eq:t0_sinc}.
    \note{
      The data were collected in
      \texttt{data/tab1} by running
      the program \texttt{invt}.
    }%
  }
  \setlength{\tabcolsep}{5pt}
  \begin{tabular} { l | p{2cm} p{2.5cm} p{1.8cm} }
    \hline
      Schedule
    &
      Inverse-time, Eq. \eqref{eq:alpha_invt1}
    &
      Modified inverse-time,
      Eq.~\eqref{eq:alpha_invtlambda}
      with optimized $\lambda$
    &
      Optimal,
      Eq. \eqref{eq:q_opt}
    \\
    \hline
    Global
    &
    $4.36(9) \times 10^{-6}$
    &
    $7.1(1) \times 10^{-7}$
    &
    $2.6(1) \times 10^{-7}$
    \\
    \hline
    Local
    &
    $3.83(7) \times 10^{-4}$
    &
    $1.72(4) \times 10^{-4}$
    &
    $9.1(2) \times 10^{-5}$
    \\
    \hline
  \end{tabular}
\end{table}

@@

Our method is most useful for long simulations to achieve relatively precise answers.
%
This allows us to assume the sampling process
to be a quasi-equilibrium one\cite{barducci2008, dama2014},
and to represent the autocorrelation function
of each eigenmode by a single number, $\Gamma_k$.
%
Although these assumptions appeared to work well
on our test system,
they may break down for relatively short simulations
of glassy systems.  Our immediate interest is to apply these relations to
problems containing aqueous mixtures of proteins and nucleic acids.

>>>>

To proceed, we first define an average\cite{
dama2014},
%
\begin{equation}
  \bar v(t)
  = \sum_{i = 1}^n \rho_i \, v_i(t)
\label{eq:vbar_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$.
%
This distribution $\pmb \rho$
will be used to measure the error,
and is independent of
the target distribution $\mathbf p$.
%
For example,
in well-tempered metadynamics,
$\pmb\rho$ is always flat,
while $\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
barducci2008, dama2014}.
%
Then, we expect the difference %between $v_i$ and $\bar v$,
%
\begin{equation}
  x_i \equiv v_i - \bar v
  %=
  %v_i - \sum_{j = 1}^n \rho_j \, v_j
  ,
\label{eq:x_def}
\end{equation}
%
to be small in the asymptotic regime.
%
It follows that
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, x_i = 0
  ,
\label{eq:rhox_sum}
\end{equation}
%
and
%
\begin{equation}
  \frac{ d \bar v (t) }
       { d t }
  =
  \alpha(t)
  \sum_{ r = 1 }^n \frac{ h_r(t) } { p_r }
  .
\label{eq:dvbardt}
\end{equation}
%
Expanding Eq. \eqref{eq:h_ave} in $x_i$ yields
%
\begin{align}
  \frac{
    \langle h_i(t) \rangle
  }
  {
    p_i
  }
  &\approx
  \frac{                        e^{- x_i} }
       { \sum_{ j = 1}^n p_j \, e^{- x_j} }
  \approx
  1 - x_i + \sum_{r = 1}^n p_r \, x_r
  ,
\label{eq:hdet_linearize}
\end{align}
%
where we have used $\sum_{j=1}^n p_j = 1$.
%
\note{For \eqref{eq:hdet_linearize}. Following the second step
$$
\langle h_i(t) \rangle
\approx
\frac{                  p_i (1 - x_i) }
     { \sum_{ r = 1 }^n p_r (1 - x_r) }
=
\frac{                      p_i (1 - x_i) }
     { 1 - \sum_{ r = 1 }^n p_r \, x_r }
.
$$
}%
From Eqs.
\eqref{eq:vt_diffeq},
\eqref{eq:h_split},
\eqref{eq:x_def}-\eqref{eq:hdet_linearize},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_WL}
\end{equation}
where
\begin{equation}
  \xi_i(t)
  =
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ r = 1 }^n
    \rho_r
    \frac{ \zeta_r(t) }
         {     p_r    }
  .
\label{eq:xi_def}
\end{equation}
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:dvbardt}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ r = 1 }^n
      \rho_r \frac{ h_r } { p_r }
  \right)
  \\
  &\stackrel{\mathrm{Eq.\; \eqref{eq:h_split}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ r = 1 }^n
      \rho_r \frac{ \langle h_r \rangle } { p_r }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:rhox_sum} \;and\; \eqref{eq:hdet_linearize}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

We define
the total error of the bias potential as
%
\begin{equation}
  \Err
  =
  \sum_{i = 1}^n \rho_i \, \langle x_i^2 \rangle
  =
  \sum_{i = 1}^n \rho_i \, \left\langle (v_i - \bar v)^2 \right\rangle
  .
\label{eq:error_sum}
\end{equation}
%
Our aim is to find the $\alpha(t)$
that minimizes the error $E$
at the end of the simulation period $T$.
%
It suffices to consider
a simple one-variable problem.

@@

To simplify the ensuing discussion,
we  limit ourselves to $\mathbf w$'s
that satisfy
the detailed balance condition,
%
\begin{equation}
  \rho_i \, w_{ij} = \rho_j \, w_{ji}
  ,
  \label{eq:w_detailedbalance}
\end{equation}
%
for the distribution $\pmb \rho$
introduced in Eq. \eqref{eq:vbar_def}.
%
It follows that
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, w_{ij}
  =
  \sum_{i = 1}^n \rho_j \, w_{ji}
  =
  \rho_j,
  \label{eq:w_balance}
\end{equation}
%
i.e., $\mathbf \rho$ is a left eigenvector of
$\mathbf w$ with eigenvalue $1$.

@@

In Fig. \ref{fig:sinc},
we give examples of the updating kernels
of the bandpass updating schemes,
compared to a Gaussian
with a similar width, $\sigma$,
%
\begin{equation}
  \sigma
  \approx
  \frac
  {
    n
  }
  {
    \sqrt{ 2 \, \pi } \, g \, K
  }
  .
\label{eq:sigma_equiv}
\end{equation}
%
%where $g = 2$ for the periodic case
%or $g = 1$ otherwise.
%
The kernel in the non-periodic case is a zigzag
because of the term $(-1)^{n+K+l-1}$
in Eq. \eqref{eq:mu_sinc_refl}.
%
However, as shown in the inset, % of Fig. \ref{fig:sinc}(b),
the ruggedness disappears
in the updating matrix by Eq. \eqref{eq:w_band_refl},
as the above oscillatory term for $l = i - j$
is canceled by the term for either $l = i + j - 1$
or $l = i + j - 2 \, n - 1$
with the opposite sign.

@@

Our test system is a one-dimensional model system, in which
$i$ represents a non-periodic variable,
and $n = 100$.
%
Both the target and intrinsic distributions
are assumed to be flat:
$p^*_i = p_i = 1/n$,
so is the distribution $\rho_i$ used in Eqs.
\eqref{eq:vbar_def} and \eqref{eq:error_sum}.
%
%We wish to see how
%the schedule, $\alpha(t)$,
%affects the fluctuating error
%of the bias potential
%at the end of the simulation.
%
Initially,
we set the bias potential to zero,
$V_i \equiv 0$,
and equilibrate the system for $10^7$ steps
under a constant updating magnitude,
$\alpha_0 = 10^{-4}$.
%
At the end of the equilibration,
the bias potential would deviate from the correct value
and our aim to see how well
the bias potential is restored
at the end of the testing schedule.
%
After the equilibration,
we reset the origin of the time, $t$, to zero,
and start the testing schedule
for $T = 10^8$ steps.
%
We repeat the above process $1000$ times (unless specified otherwise)
and average the results.
%specified by either
%Eq. \eqref{eq:q_opt}
%or
%Eq. \eqref{eq:alpha_invtlambda}.

@@

\begin{table}[h]
  \caption{\label{tab:error_Gaussian}
    Error of the Gaussian updating scheme of $\sigma = 10$.
    $t_0$ is given by Eq. \eqref{eq:t0_sinc}.
    \note{
      The data were collected in
      \texttt{data/tab1} by running
      the program \texttt{invt}.
    }%
  }
  \setlength{\tabcolsep}{2pt}
  \renewcommand\arraystretch{1.2}
  \begin{tabular} { l | p{2.2cm} p{2.4cm} p{2.0cm} }
    \hline
      Schedule
    &
      Inverse-time, Eq. \eqref{eq:alpha_invt1}
    &
      Modified inverse-time,
      Eq.~\eqref{eq:alpha_invtlambda}
      with optimized $\lambda$
    &
      Optimal,
      Eq. \eqref{eq:q_opt}
    \\
    \hline
    Global
    &
    $4.36(9) \times 10^{-6}$
    &
    $7.1(1) \times 10^{-7}$
    &
    $2.6(1) \times 10^{-7}$
    \\
    \hline
    Local
    &
    $3.83(7) \times 10^{-4}$
    &
    $1.72(4) \times 10^{-4}$
    &
    $9.1(2) \times 10^{-5}$
    \\
    \hline
  \end{tabular}
\end{table}

@@

Our method is most useful for long simulations to achieve relatively precise answers.
%
This allows us to assume the sampling process
to be a quasi-equilibrium one\cite{barducci2008, dama2014},
and to represent the autocorrelation function
of each eigenmode by a single number, $\Gamma_k$.
%
Although these assumptions appeared to work well
on our test system,
they may break down for relatively short simulations
of glassy systems.
%Our immediate interest is to apply these relations to
%problems containing aqueous mixtures of proteins and nucleic acids.
